---
title: 'Project 1: Exploratory Data Analysis'
author: Tina Gray ttg333
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

## Data Wrangling and Data Exploration


```{r}
library(tidyverse)
library(readr)
X2016_tooth_decay_Sheet1 <- read_csv("2016 tooth decay - Sheet1.csv")
library(readr)
X2016_drug_overdose_data_Sheet1 <- read_csv("2016 drug overdose data - Sheet1.csv")
library(readr)
Current_Cigarette_Use_Among_Adults_Behavior_Risk_Factor_Surveillance_System_2018 <- read_csv("Current Cigarette Use Among Adults (Behavior Risk Factor Surveillance System) 2018.csv")
cigs <- Current_Cigarette_Use_Among_Adults_Behavior_Risk_Factor_Surveillance_System_2018
drugs <- X2016_drug_overdose_data_Sheet1
teeth <- X2016_tooth_decay_Sheet1
cigs <- cigs %>% rename(cig.use.rate = Data_Value, cig.n = Sample_Size)
drugs <- drugs %>% rename(od.death.rate = `Age Adjusted Rate`, n.deaths = `Number of Deaths`)
teeth <- teeth %>% rename(lost.teeth = `Percent of Seniors Who Have Lost Teeth`)
#renaming my data because the names are very long, renaming columns so they make more sense after combining the datasets.
```
I chose three datasets. The teeth dataset contains the percentage of seniors (65+) who have lost some of their teeth by state. The cigs dataset contains the percentage of adults who smoke and the sample size by state. The drugs dataset contains opioid overdose death rates and the number of deaths by state. I chose these datasets because I am interested in going to dental school in the future. Cigarettes and drug use are both known to damage the health of teeth, so I chose those datasets to compare to the teeth data to see if there was a correlation between the usage of those things and the health of the teeth of seniors by state. I think that states with higher tooth loss will also tend to have higher rates of smoking and drug overdose.




```{r}
teethanddrugs <- full_join(teeth,drugs)
fulldata <- left_join(teethanddrugs,cigs)
glimpse(fulldata)
```
There were 53 observations in the cigs dataset and 50 observations in both the teeth and drugs data. I did a full join for teeth and drugs because they had data for each state and it would've given me the same result regardless of which type of join I did. I did a left join to join that data with the cigs data because the cigs data had a few extra observations that I wanted to exclude, since they weren't in the other datasets. The cases that were dropped out of the cigs dataset were data collected from the U.S territories and District of Columbia. Dropping those cases should not be an issue since only 3 cases were dropped.

    
## Summary Stats
```{r}
fulldata <- fulldata %>% na.omit
fulldata %>% filter(lost.teeth>20) %>% arrange(desc(lost.teeth))
fulldata %>% summarise_if(is.numeric, mean)
fulldata %>% summarize_if(is.numeric, sd)
fulldata %>% summarize_if(is.numeric, max)
fulldata %>% summarize_if(is.numeric, min)
fulldata %>% summarize_if(is.numeric, n_distinct)
fulldata %>% select_if(is.numeric) %>% cor()
fulldata <- fulldata %>% mutate(abuse.cat = ifelse(od.death.rate > 20, "high",
                                       ifelse(od.death.rate <=20 & od.death.rate<=10, "med", "low")))
fulldata %>% group_by(abuse.cat,Location) %>% summarize(drugs.n = (1/od.death.rate)*n.deaths)
#drugs.n gives the total number of overdoses in the sample, including those that did not result in death
fulldata %>% group_by(abuse.cat) %>% summarize(mean_od = mean(od.death.rate), mean_smoke =
                                                 mean(cig.use.rate), mean_lost_teeth = 
                                                 mean(lost.teeth)) %>% 
  pivot_longer(contains ("_")) %>% separate(name, into = c("function", "type")) %>% 
  pivot_wider(names_from = type, values_from = value)

```
 The first thing I did was omit any NAs so that none of my summary statistics would be messed up. Then, I filtered the rows with more than 20% lost teeth and arranged them to see the states with the largest percentage of seniors who have lost teeth. After that, I used summarize if to find the mean, standard deviation, minimum, maximum, and number of distinct values for each column. Then, I mutated by data to add a new column, abuse.cat. abuse.cat sorts each state into categories "high", "med", and "low" based on the drug overdose rates. I used this new column and summarize to find drugs.n, which is the total number of overdoses in the sample.
 I grouped by the abuse category to calculate the mean of overdoses, smokers, and teeth lost. I used pivot longer and pivot wider to tidy the data up. The data wasn't really untidy in the first place sinse all of the statistics were means, but the names of each column are a little cleaner since they don't have "mean_" in front of them anymore. Instead, the mean is in its own column called function. First I used pivot_longer on the numeric columns, then I separated them by the name to separate the type of summary statistic (mean) and the data that the mean function was performed on. Afterwards, I used pivot_wider to make the overdoses, smokers, and teeth lost back into column names.
 
    
## Data Visualization
```{r}
library(ggplot2)
cormat <- fulldata %>% select_if(is.numeric) %>% cor(use ="pair")
tidycor <- cormat %>% as.data.frame %>% rownames_to_column("var1") %>%  pivot_longer(-1,names_to="var2",values_to="correlation")
tidycor%>%ggplot(aes(var1,var2,fill=correlation))+  geom_tile()+  scale_fill_gradient2(low="red",mid="white",high="blue")+  geom_text(aes(label=round(correlation,2)),color = "black", size = 4)+ theme(axis.text.x = element_text(angle = 90, hjust=1))+ coord_fixed() + ggtitle("Correlation Heatmap")
```
The correlation heatmap shows a strong correlation between lost teeth and cigarette use rate. All other rates are much lower. n.deaths and cig.n have the next highest correlation, but the relationship between them makes less sense because n.cigs is a sample size for the cigarette data and has little to do with the number of deaths from opioid overdose.
```{r}
ggplot(data = fulldata, aes(x = lost.teeth, y= cig.use.rate, color = abuse.cat)) +
         geom_point() +theme_minimal()+ggtitle("Teeth Lost vs. Cigarette Use by State")+
  xlab("Percentage of Seniors Who Have Lost Teeth") + ylab("Cigarette Use Rate") + scale_color_manual(breaks = c("high", "med", "low"),
                        values=c("red", "yellow", "green"))
```
Since the correlation heatmap showed that lost.teeth and cig.use.rate had a strong correlation, I wanted to see a scatterplot of the two. I colored the points by abuse.cat to incorporate the drug data into this plot as well. This plot confirms that there is a strong correlation between the two variables because you can easily see an upward linear trend.
```{r}
fulldata <- fulldata %>% mutate(smoke.cat = ifelse(cig.use.rate > 20, "high smoke",
                                       ifelse(cig.use.rate <=20 & cig.use.rate<=10, "med smoke", "low smoke")))
ggplot(fulldata, aes(x = abuse.cat,y = od.death.rate, fill = abuse.cat)) +geom_bar(stat = "summary") + geom_errorbar(stat= "summary", width =.5) +theme_dark()+scale_fill_brewer()+
  facet_wrap(~smoke.cat) + scale_y_continuous(name = "Overdose Death Rate") +
  scale_x_discrete("Abuse Catergory")+ggtitle("Overdose Death Rate Breakdown")
```
For the bar graphs, I mutated a new categorical column in my data called smoke.cat. smoke.cat is the same as abuse.cat, but instead of being based on drug overdose rates, it is based on cigarette smoking rates. I made this new column to facet wrap my bar graph (so that I could incorporate 3 variables) without having to facet wrap it by state. It would have been too overwhelming to do it by state and it would be really hard to read that way. These bar graphs show the overdose death rates for each abuse category, while being separated by high, medium, and low smoking rates.
    
## Clustering     
```{r}
library(cluster)
pam_dat<-fulldata%>%select(lost.teeth,od.death.rate,n.deaths,cig.use.rate,cig.n)
sil_width<-vector()
for(i in 2:10){  
  pam_fit <- pam(pam_dat, k = i)  
  sil_width[i] <- pam_fit$silinfo$avg.width  
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)
pam1 <- fulldata %>% select_if(is.numeric) %>% scale %>% pam(k=3)
pam1
```
For this code, I made a plot to figure out the number of clusters to use for PAM. 3 has the highest sil_width, so 3 is best choice for the number of clusters. I did PAM in this section of the code as well.
```{r}
eig1 <- fulldata %>% select_if(is.numeric) %>% cor() %>% eigen()
x <- fulldata %>% select_if(is.numeric) %>% scale
PCAscores <- x%*%eig1$vectors
fulldata%>%mutate(PC1=PCAscores[,1], PC2=PCAscores[,2])%>%
  ggplot(aes(PC1,PC2, color = pam1$clustering))+geom_point()
```
I made eigenvectors for the data to do PCA. I made a matrix of the numeric data in my fulldataset and multiplied it with the eigenvectors. I made a plot to visualize my clusters, and colored them by the pam clusters.
```{r}
plot(pam1, which=2)
```
This plot shows that the average silhouette width is .14. Since the width is less than .25, no substantial structure has been found. This is supported by the plot above the silhouette wideth plot, where the clusters were all randomly scattered around the graph and did not appear to cluster anywhere. 
<P style="page-break-before: always">
\newpage
    
